---
title: "Stock Dynamics"
author: "By John Bobo based on a problem set from MIT's Analytics Edge MOOC" 
date: "April 23, 2016"
output:
    html_document:
        theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
A stock market is where buyers and sellers trade shares of a company, and is one of the most popular ways for individuals and companies to invest money. The size of the world stock market  is now estimated to be in the trillions. The largest stock market in the world is the New York Stock Exchange (NYSE), located in New York City. About 2,800 companies are listed on the NSYE. In this problem, we'll look at the monthly stock prices of five of these companies: [IBM](http://www.ibm.com/), [General Electric (GE)](http://www.ge.com/), [Procter and Gamble](http://www.pg.com/), [Coca Cola](http://www.coca-cola.com/), and [Boeing](http://www.boeing.com/). The data used in this problem comes from [Infochimps](http://www.infochimps.com/).

Download and read the following files into R, using the read.csv function: [IBMStock.csv](https://d37djvu3ytnwxt.cloudfront.net/asset-v1:MITx+15.071x_3+1T2016+type@asset+block/IBMStock.csv), [GEStock.csv](https://d37djvu3ytnwxt.cloudfront.net/asset-v1:MITx+15.071x_3+1T2016+type@asset+block/GEStock.csv), [ProcterGambleStock.csv](https://d37djvu3ytnwxt.cloudfront.net/asset-v1:MITx+15.071x_3+1T2016+type@asset+block/ProcterGambleStock.csv), [CocaColaStock.csv](https://d37djvu3ytnwxt.cloudfront.net/asset-v1:MITx+15.071x_3+1T2016+type@asset+block/CocaColaStock.csv), and [BoeingStock.csv](https://d37djvu3ytnwxt.cloudfront.net/asset-v1:MITx+15.071x_3+1T2016+type@asset+block/BoeingStock.csv). (Do not open these files in any spreadsheet software before completing this problem because it might change the format of the Date field.)

Call the data frames "ibm", "ge", "pg", "coke", and "boeing", respectively. Each data frame has two variables, described as follows:

* **Date**: the date of the stock price, always given as the first of the month.  
* **StockPrice**: the average stock price of the cmopany in the given month.
```{r}
ibm <- read.csv("/Users/johnbobo/analytics_edge/data/IBMStock.csv")
ge <- read.csv("/Users/johnbobo/analytics_edge/data/GEStock.csv")
pg <- read.csv("/Users/johnbobo/analytics_edge/data/ProcterGambleStock.csv")
coke <- read.csv("/Users/johnbobo/analytics_edge/data/CocaColaStock.csv")
boeing <- read.csv("/Users/johnbobo/analytics_edge/data/BoeingStock.csv")
```

We'll take a look at how the stock dynamics of these companies have changed over time.

## Preliminary Concerns
Before working with these data sets, we need to convert the dates into a format that R can understand. Take a look at the structure of one of the datasets using the `str` function:
```{r}
str(ibm)
```

Right now, the date variable is stored as a factor. We can convert this to a "Date" object in R by using the following five commands (one for each data set):
```{r}
ibm$Date = as.Date(ibm$Date, "%m/%d/%y")
ge$Date = as.Date(ibm$Date, "%m/%d/%y")
coke$Date = as.Date(ibm$Date, "%m/%d/%y")
pg$Date = as.Date(ibm$Date, "%m/%d/%y")
boeing$Date = as.Date(ibm$Date, "%m/%d/%y")
```
The first argument to the as.Date function is the variable we want to convert, and the second argument is the format of the Date variable. We can just overwrite the original Date variable values with the output of this function. Now, we'll answer the following questions using the str and summary functions.

## Summary Statistics
**Our five datasets all have the same number of observations. How many observations are there in each data set?**
```{r}
nrow(ge)
```
**Answer**: 480

***

**What is the earliest year in our datasets?**
```{r}
min(ge$Date[1])
```
**Answer**: 1970

***

**What is the latest year in our datasets?**
```{r}
max(ge$Date)
```
**Answer**: 2009

***

**What is the mean stock price of IBM over this period of time?**
```{r}
mean(ibm$StockPrice)
```
**Answer**: 144.375

***

**What is the minimum stock price of General Electric (GE) over this time period?**
```{r}
min(ge$StockPrice)
```
**Answer**: 9.294

***

**What is the maximum stock price of Coca-Cola over this time period?**
```{r}
max(coke$StockPrice)
```
**Answer**: 146.584

***

**What is the median stock price of Boeing over this time period?**
```{r}
median(boeing$StockPrice)
```
**Answer**: 44.883

***

**What is the standard deviation of the stock price of Procter & Gamble over this time period?**
```{r}
sd(pg$StockPrice)
```
**Answer**: 18.194

***

**Pro-tip**: We can most of these answers using the `summary` function:
```{r}
summary(ibm)
```

***

## Visualizing Stock Dynamics
Let's plot the stock prices to see if we can visualize trends in stock prices during this time period. Using the plot function, plot the Date on the x-axis and the StockPrice on the y-axis, for Coca-Cola.

This plots our observations as points, but we would really like to see a line instead, since this is a continuous time period.
```{r warning=FALSE}
library(ggplot2)
library(scales)
plot <- ggplot(coke, aes(x=Date, y=StockPrice)) + geom_line(color="red") +
            ggtitle("Stock Prices 1970-2010") +
            ylab("Price")
plot
```

We can see Coca-Cola had its highest stock price in 1973 and its lowest price in 1980.

For future ease let us make a datafram with all the prices together
```{r}
df <- cbind(ibm, ibm$StockPrice, ge$StockPrice, pg$StockPrice, coke$StockPrice, boeing$StockPrice)
df <- df[,-2]
colnames(df) <- c("Date", "ibmPrice", "gePrice", "pgPrice", "cokePrice", "boeingPrice")
```


Now let's add a line for Procter & Gamble too. 
```{r}
plot <- ggplot(df, aes(x=Date)) + 
            geom_line(aes(y=cokePrice, color="Coca-Cola")) +
            geom_line(aes(y=pgPrice, color="Procter & Gamble")) +
            ylab("Stock Price") +
            ggtitle("Stock Prices from 1970-2010") +
            scale_x_date() +
            geom_vline(aes(xintercept = unclass(as.Date("2000-03-01"))), 
                       linetype = "dashed", size=.5, alpha = .3) +
            geom_vline(aes(xintercept = unclass(as.Date("1983-01-01"))), 
                       linetype = "dashed", size=.5, alpha = .3)
            
plot
```

**In March of 2000, the technology bubble burst, and a stock market crash occurred. According to this plot, which company's stock dropped more?**  
**Answer**: Procter & Gamble.

***

**Around 1983, the stock for one of these companies (Coca-Cola or Procter and Gamble) was going up, while the other was going down. Which one was going up?**  
**Answer**: Coca-Cola.

***

## Visualizing Stock Dynamics 1995 - 2005

Let's take a look at how the stock prices changed from 1995-2005 for all five companies.
```{r}
plot <- ggplot(subset(df, 
                      Date > as.Date("1995-01-01") & Date < as.Date("2006-01-01")),
                aes(x=Date)) +
            geom_line(aes(y=cokePrice, color="Coca-Cola")) +
            geom_line(aes(y=pgPrice, color="Procter & Gamble")) +
            geom_line(aes(y=gePrice, color="GE")) +
            geom_line(aes(y=ibmPrice, color="IBM")) +
            geom_line(aes(y=boeingPrice, color="Boeing")) +
            ylab("Stock Price") +
            ggtitle("Stock Prices from 1995-2005") +
            scale_x_date() +
            geom_vline(aes(xintercept = unclass(as.Date("2000-03-01"))), 
                       linetype = "dashed", size=.5, alpha = .3) +
            geom_vline(aes(xintercept = unclass(as.Date("1997-09-01"))), 
                       linetype = "dashed", size=.5, alpha = .3) +
            geom_vline(aes(xintercept = unclass(as.Date("1997-11-01"))), 
                       linetype = "dashed", size=.5, alpha = .3)
plot      
```

**Which stock fell the most right after the technology bubble burst in March 2000?**  
**Answer**: General Electric (GE).

***

**Which stock reaches the highest value in the time period 1995-2005?**  
**Answer**:IBM

***

**In October of 1997, there was a global stock market crash that was caused by an economic crisis in Asia. Comparing September 1997 to November 1997, which companies saw a decreasing trend in their stock price?**  
**Answer**: Boeing and Procter & Gamble.

***

**In the last two years of this time period (2004 and 2005) which stock seems to be performing the best, in terms of increasing stock price?**  
**Answer**: Boeing.

***

## Monthly Trends
Lastly, let's see if stocks tend to be higher or lower during certain months. Use the tapply command to calculate the mean stock price of IBM, sorted by months.

```{r}
sort(tapply(ibm$StockPrice, months(ibm$Date), mean), decreasing = TRUE)
mean(ibm$StockPrice)
```

**For IBM, compare the monthly averages to the overall average stock price. In which months has IBM historically had a higher stock price (on average)?**  
**Answer**: January, February, March, April, and May.

***

**For the months of December and January, every company's average stock is higher in one month and lower in the other. In which month are the stock prices lower?**  
**Answer**: December.

